---
title: Event types that are most harmful with respect to population health and economics in the USA
author: "justmzen"
date: "6/01/2020"
output: 
        html_document: default
        md_document: default
---

# Synopsis
Storms and other severe weather events can cause both public health and economic 
problems for communities and municipalities. Many severe events can result in 
fatalities, injuries, and property damage, and preventing such outcomes to the
extent possible is a key concern.  

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of 
major storms and weather events in the United States, including when and where 
they occur, as well as estimates of any fatalities, injuries, and property 
damage.  

The analysis of such database shows that droughts, hails and floods are the most 
harmful event types with respect to economics, whereas tornadoes, excessive 
heat and floods are the most harmful with respect to public health. In 
particular, floods are in the top 3 in both cases.  

# Data Processing

## Data loading
The first part consists in downloading the [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
from the online repository and loading it as a data.table variable.  

```{r, cache = TRUE, echo = TRUE}

## Libraries loading
library('data.table')
library('ggplot2')

## Data loading
# Storm Data downloading
if(!file.exists('./stormData.bz2')){
        fileURL <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
        download.file(url = fileURL, destfile = './stormData.bz2')
        rm(fileURL)
        dateDownloaded <- date()
}

# Loading the data set
sd <- as.data.table(read.csv(file = './stormData.bz2'))

```

## Variables management

The variables in the data set are:  
```{r, cache = TRUE, echo = TRUE}
# Variables list
names(sd)
```

BGN_DATE is the date related to each observation, hence it is useful to specify
it as date class.

```{r, cache = TRUE, echo = TRUE}
# Data manipulation
sd$BGN_DATE <- as.Date(sd$BGN_DATE, format = '%m/%d/%Y %H:%M:%S')
```

## Data subsetting

Since the objective is comparing the effects of different weather events, the
data set has been subsetted for dates after Jan. 1996, date in which the
observatory started recording all event types.  

Furthermore, only just 8 variables out of 37 are useful to answer the initial 
question:  
- BGN_DATE: date of the event  
- EVTYPE: event type  
- FATALITIES: number of deaths  
- INJURIES: number of injuried people  
- PROPDMG: property damage as coefficient  
- PROPDMGEXP: exponent (of 10) of the property damage  
- CROPDMG: crop damage as coefficient  
- CROPDMGEXP: exponent (of 10) of the crop damage  

```{r, cache = TRUE, echo = TRUE}
# Data set subsetting
# Filtering by date
sd <- sd[BGN_DATE >= as.Date('01/01/1996', format = '%m/%d/%Y')]

# Choosing useful variables
sd <- sd[, .(BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP,
             CROPDMG, CROPDMGEXP)]
```

## Conversion of values in PROPDMGEXP and CROPDMGEXP variables

The mentioned variables present observations which are both numerical and
alphabetical.  
The work entitled ["How To Handle Exponent Value of PROPDMGEXP and CROPDMGEXP"](https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html) discusses the issue in depth.  
The following values found in the two variables will be converted in:  
- h/H: 100  
- k/K: 1,000  
- m/M: 1,000,000  
- b/B: 1,000,000,000  
- +: 10  
- -, ? and blank: 0  

```{r, cache = TRUE, echo = TRUE}
# Managing EXP values
conversionValues <- c('0' = 10^0,
                      '1' = 10^1,
                      '2' = 10^2,
                      '3' = 10^3,
                      '4' = 10^4,
                      '5' = 10^5,
                      '6' = 10^6,
                      '7' = 10^7,
                      '8' = 10^8,
                      '9' = 10^9,
                      '+' = 10^1,
                      '-' = 10^0,
                      '?' = 10^0,
                      'h' = 10^2, 'H' = 10^2,
                      'k' = 10^3, 'K' = 10^3,
                      'm' = 10^6, 'M' = 10^6,
                      'b' = 10^9, 'B' = 10^9
)

# Conversion
sd <- sd[, PROPDMGEXP := conversionValues[sd[, PROPDMGEXP]]]
sd <- sd[is.na(PROPDMGEXP), PROPDMGEXP := 10^0]
sd <- sd[, CROPDMGEXP := conversionValues[sd[, CROPDMGEXP]]]
sd <- sd[is.na(CROPDMGEXP), CROPDMGEXP := 10^0]
rm(conversionValues)
```

Finally, the real value of damages to properties and crops are evaluated by
the moltiplication between the coefficient (i.e. PROPDMG and CROPDMG) and the
relative exponent (i.e. PROPDMGEXP and CROPDMGEXP).  
The new variables are named PROPDMGTOT and CROPDMGTOT, respectively, and replace
the four variables used to calculate them.

```{r, cache = TRUE, echo = TRUE}
# Replacing PROP and CROP columns
sd <- sd[, .(BGN_DATE, EVTYPE, FATALITIES, INJURIES,
             PROPDMGTOT = PROPDMG*PROPDMGEXP,
             CROPDMGTOT = CROPDMG*CROPDMGEXP)]
```

## Global calculations by event type

The global damages are evaluated by event type and are calculated as the sum of
the damages to properties and crops. Only the first 15 will be considered.  

```{r, cache = TRUE, echo = TRUE}
# Damage evaluation: sum of damages by event type
damageToBelongings <- sd[, .(damage = sum(PROPDMGTOT) + sum(CROPDMGTOT)),
                         by = EVTYPE]
damageToBelongings <- damageToBelongings[order(damage, decreasing = TRUE), ]
damageToBelongings <- damageToBelongings[1:15, ]
damageToBelongings$EVTYPE <- droplevels(damageToBelongings$EVTYPE)
```

The same thing is applied to the people involved, calculated as the sum of
injuried and dead people.

```{r, cache = TRUE, echo = TRUE}
# Injuries and deaths by event type
damageToPeople <- sd[,.(injuries = sum(INJURIES),
                        fatalities = sum(FATALITIES),
                        peopleInvolved = sum(INJURIES) + sum(FATALITIES)),
                     by = EVTYPE]
damageToPeople <- damageToPeople[order(peopleInvolved, decreasing = TRUE), ]
damageToPeople <- damageToPeople[1:15, ]
damageToPeople$EVTYPE <- droplevels(damageToPeople$EVTYPE)
```

# Results

The most harmful event types on properties and crops are plotted:  
```{r, cache = TRUE, echo = TRUE}
# Total damage on properties and crops
ggplot(data = damageToBelongings, 
       aes(x = reorder(EVTYPE, -damage), y = damage)) +
        geom_bar(stat = 'identity') + 
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(title = 'Total damage in USD from 1996 by event type',
             x = 'Event type', y = 'Total damage [USD]')
```

The bar plot shows that the 6 most harmful event types on properties and crops
are:  

```{r, cache = TRUE, echo = TRUE}
head(damageToBelongings)
```

Whereas, the most harmful event types on people are:

```{r, cache = TRUE, echo = TRUE}
# Total people involved
head(damageToPeople)

ggplot(data = damageToPeople,
       aes(x = reorder(EVTYPE, -peopleInvolved), y = peopleInvolved)) +
        geom_bar(stat = 'identity') + 
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(title = 'Total number of people involved from 1996 by event type',
             x = 'Event type', y = 'People')
```
